Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2024-04-27 report/manuscript.tex
rearrange markdown sections

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,floatsintext]{apa7}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
  pdflang={en-EN},
  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Modelling writing hesitations}
\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{colortbl}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage[utf8]{inputenc}
\usepackage{icomma}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

@@

\abstract{
Classical serial models view the process of producing a text as a descrete chain of pauses and writing bursts. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two view in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found that although pause durations are consistently longer for larger linguistic edges, the pause frequencies are not, but largely identical for before sentence and before word transition locations. Our results cannot be explained by the serial but are in line with the cascading view of writing.
}

@@

@@

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. Rønneberg et al. (2022) coined this possibility the \enquote{process-disruption hypothesis.}

@@

We (re)analysed keystroke data of 6 experiments in which participants composed text in a series of 5 Bayesian models. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models. In particular we used a Gaussian distribution, a log-normal model, and an unequal variance model that takes into consideration that longer latencies -- those associated with larger linguistic edges -- are known to be associated with a larger variance component (Schöner, 2002; Wagenmakers \& Brown, 2007; Wing \& Kristofferson, 1973).

The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in Roeser et al. (2021). Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow. Importantly these two models assume that hesitations are not sufficiently determined by transition location -- as the serial account does -- but also transition locations are associated with different probabilities that hesitation may occur. This view was implemented as a constrained and an unconstrained model.

We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts. We discussed advantages with this approach in the introduction section. These datasets include samples from populations with different writing experience and languages (e.g.~young / L2 writers, students) and writing tasks (e.g.~essays, syntheses). Datasets from a variety of different writers and writing tasks were deliberately chosen to challange our modelling approach and to test to what extent resulting pausing patterns generalise across writing contexts.

@@

Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; details will be presented below.

@@

\begin{tabular}{llllrrp{2cm}p{3cm}}
\toprule
Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N (ppts)} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Language} & \multicolumn{1}{c}{Manipulation}\\
\midrule
C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.80 & Norwegian & --\\
CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.90 & Norwegian & weak decoders / control; masked / unmasked\\
GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite &  & 45 & NA & Norwegian & masked / unmasked\\
LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 16.95 & Dutch & Various topics and genres\\
PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23.00 & English (L2) & pre / post test trained in plain language principles and control\\
SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.60 & English (L1) / Spanish (L2) & write in L1 / L2\\
\bottomrule
\end{tabular}

@@

\end{table}
\end{landscape}

TODO: Mark, can you add information for C2L1 and CATO as needed

TODO: might need to remove kids that don't speak Norwegian at home (see github issue)? I don't think that's necessary though.

For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012).

The C2L1 dataset comprises data Norwegian 6th graders -- \emph{N} = 126, mean age 11 years 10 months -- published in Rønneberg et al. (2022). The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.

The CATO dataset is published in Torrance, Rønneberg, et al. (2016). Norwegian upper secondary students -- \emph{N} = 26, mean age = 16.9 years -- with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.

TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say \enquote{unpublished.} I could also just use Escop as reference (also for SPL2; maybe my SIG Writing talk). Was the sample Norwegian?

The GUNNEXP2 dataset published in Torrance and Ofstad (n.d.) includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX). In this dataset participants wrote texts either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal.

The SPL2 data set involves undergraduate university students in the US -- \emph{N} = 39, 2mean age = 20.6 years, SD = 1.51 -- composing argumentative essays in English (L1) and Spanish (L2). Students wrote within the CyWrite word-processing environment (Chukharev-Hudilainen et al., 2019). Participants wrote, with a 40 minute time limit, to two writing prompts. Order and writing prompt were counterbalanced across language.

LIFT and PLanTra data data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021).

The LIFT data are published in Vandermeulen, Steendam, et al. (2020) and described in Vandermeulen, De Maeyer, et al. (2020). LIFT is a representative sample of Dutch students (\emph{N} = 658, mean age = 16.95 years, 428 females and 230 males) in the three highest grades of pre-university education collected from 43 schools. The students were asked to conduct four synthesis writing tasks, two argumentative and two informative texts, about each of four topics, with order randomized per school. The students received 50 minutes for each tasks.

The PLanTra (Plain Language Training for business content) data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a). The primary aim of the project was to investigate the impact of plain language instruction on business students' strategies to simplify business texts as well as on the comprehensibility of the produced texts. A total of \emph{N} = 47 graduate students (mean age = 23 years, 38 females and 9 males, 45 native Dutch speakers) of the master Business and Economics participated. The study adopted a pre-test post-test design. As pre-test, participants were asked to rewrite a given text (extract of a corporate report on sustainability), to make it more engaging and easier to read for a lay audience. Thereafter the experimental group received online instruction on how to apply plain language principles to sustainability content, while the control group received online instruction exclusively on the topic of sustainability. Participants were asked to spend at least 45 minutes on the instruction module. As post-test, participants were asked 2-3 days later to simplify another extract of a corporate report on sustainability. Both reports were written in English (second language) and similar in length (274-278 words) and readability. Participants received as much time as needed for each task.

TODO: Mark, can you add information here?

The SPL2 dataset is published in Torrance et al. (n.d.). The data come from a text composition task of undergraduate university students -- \emph{N} = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite (Chukharev-Hudilainen et al., 2019). Participants wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.

\hypertarget{transition-types}{%
\subsection{Transition types}\label{transition-types}}

The transition types that were analysed in this study focus on those locations that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance et al., n.d.; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}. In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions. At before-sentence locations, IKIs were timed to the shift keypress for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we will return to this issue in the Results section. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis.

@@

\begin{tabular}{p{3cm}p{3cm}p{8cm}}
\toprule
Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
\midrule
Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d. T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
\bottomrule
\addlinespace
\end{tabular}

@@

\hypertarget{data-reduction}{%
\subsection{Data reduction}\label{data-reduction}}

For all datasets we only used transitions that were not followed by an editing operation. Transitions that occurred at the beginning of the text or the beginning of a paragraph were removed. We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 in the LIFT data set, and 41 participants in the PLanTra data set). We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant).

We further removed keystroke intervals that are extremely short (\(\le\) 50 msecs) or extremely long (\(\ge\) 30 secs); the percentage of remove keystroke data can be found in Table \ref{tab:datareduction}. From the remaining data we randomly sampled 100 observations per participant, per condition, and per transition location, with the exception of the LIFT data set. This was done for computational reasons to reduce the time the Bayesian models -- decribed in the next section -- require to complete sampling. For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our data pool. Because the LIFT data set included the large number of writing tasks as fixed effect, we sampled 50 observations per condition, location and participant to not exceed our computational resources. The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.

@@

\caption{\label{tab:datareduction}Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

@@

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Keystroke modelling; finite mixture models; Bayesian models; text composition}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{colortbl}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage[utf8]{inputenc}
\usepackage{icomma}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{csquotes}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process},
  pdfauthor={Jens Roeser1, Mark Torrance1, Rianne Conijn2, \& Evgeny Chukharev3},
  pdflang={en-EN},
  pdfkeywords={Keystroke modelling; finite mixture models; Bayesian models; text composition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\shorttitle{Modelling writing hesitations}

@@

\abstract{%
Classical serial models view the process of producing a text as a discrete chain of pauses and writing bursts. In contrast, parallel cascading models of writing assume that planning is not complete at production onset and operates in parallel to writing execution. We implemented these two views in Bayesian statistical models and applied our models to key-stroke logs of 6 data sets from free text production. We reanalysed keystroke intervals at before-sentence transitions, before word transitions and within word transitions. Model comparisons demonstrated strong evidence in favour of the statistical implementation of the cascading view of the serial models across all data sets. Further we found -- across populations -- that transitions across word and even sentence boundaries are often fluent and uninterrupted by planning of upcoming linguistic material; when pausing occurred pauses were longer before sentence than before word which is consistent with the idea larger linguistic edges are associated with higher level planning. Our results cannot be explained by the serial but are in line with the cascading view of writing.
}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what Christiansen and Chater (2016) describe as a fundamental \enquote{now-or-never} bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. Rønneberg et al. (2022) coined this possibility the \enquote{process-disruption hypothesis}.

@@

In a series of five Bayesian models, we (re)analysed keystroke data form six experiments in which participants composed texts. Full modelling details can be found in Appendix \textbackslash ref\{statistical-models; we summarise the most important model properties in the following. The first three models map onto the serial account of writing hesitations and bridge between commonly used statistical models for the analysis of inter-keystroke intervals (IKI) as single distribution models and a principled single-distribution model of keystroke data. In particular we used a Gaussian distribution, and a log-Gaussian model, similar to statistical models that can frequently be found in the literature, and an unequal variance model that takes into consideration that intervals associated with larger linguistic edges may have a larger variability.

The remaining two models are implementations of the cascading view following the two-distributions mixture-model approach presented in Roeser et al. (2021). Importantly these two models assume that IKIs result from a mixture of two processes: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels cause delays in the information flow and therefore lead to longer lags between keystrokes. These two models assume that hesitations are not sufficiently determined by transition location -- as in the serial account -- but transition locations are also associated with different probabilities that a hesitation occurs. This view was implemented as a constrained model that assumes no keystroke-interval difference between transition locations for uninhibited activation flow and an unconstrained model which assumes that keystroke-transitions may vary between transition locations even when information flow from higher to lower levels is uninterrupted.

We reanalysed datasets from six experiments in which researchers elicited full texts in response to writing prompts. We discussed advantages with this approach in the introduction section. These datasets include samples from populations with various writing experience and languages (e.g.~young / L2 writers, students) performing different writing tasks (e.g.~essays, syntheses). Datasets from a variety of different populations of writers and writing tasks were deliberately chosen to challenge our modelling approach and to test to what extent pausing patterns generalise across writing contexts.

@@

TODO: Mark, can you check / add information for C2L1, CATO, GUNNEXP2. For GUNNEXP2 I've added a provisional citation that probably needs changing; we could just say \enquote{unpublished}. I could also just use the Escop presentation as reference (also for SPL2; maybe my SIG Writing talk). Was the GUNNEXP2 sample Norwegian? Should we include information like time limit or text length in the table?

Six datasets with keystroke data from free text production were used for analysis. An overview can be found in Table \ref{tab:datasets}; descriptions below. Dataset used are C2l1 (Rønneberg et al., 2022), CATO (Torrance, Rønneberg, et al., 2016), GUNNEXP2 (Torrance \& Ofstad, n.d.), SPL2 (Torrance et al., n.d.), LIFT published in Vandermeulen, Steendam, et al. (2020) and described in Vandermeulen, De Maeyer, et al. (2020), PLanTra data are published in Rossetti and Van Waes (2022b) and described in Rossetti and Van Waes (2022a). In C2L1 Norwegian 6th graders composed argumentative essays. In CATO upper secondary students in Norway with weak decoding skills (and control) composed expository texts either normally or with masked letters to prevent reading their unfolding text. In SPL2 undergraduate students produced argumentative essays in their L1 (English) or L2 (Spanish); order of language and two writing prompts were counterbalanced. The LIFT data contains pre-university students producing argumentative and informative text syntheses on four topics each. The PLanTra data contains data from Master students in Business and Economics simplifying texts on sustainability before and after received either online instruction on how to apply plain language principles to sustainability content or a online instruction exclusively on the topic of sustainability. The GUNNEXP2 data \ldots. {[}MARK, please add a sentence here to describe the production context{]}.

For C2L1, CATO, and GUNNEXP2, keystroke data were captured using EyeWrite (Simpson \& Torrance, 2007; Torrance, 2012). LIFT and PLanTra data were captured using InputLog (Leijten \& Van Waes, 2013; Van Waes et al., 2019, 2021) and SPL2 data were collected using CyWrite (Chukharev-Hudilainen et al., 2019).

@@

\begin{tabular}{lp{3cm}p{2cm}p{2.5cm}rrp{2.5cm}p{2.25cm}p{2.5cm}}
\toprule
Dataset & \multicolumn{1}{c}{Source} & \multicolumn{1}{c}{Keylogger} & \multicolumn{1}{c}{Writing task} & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean age} & \multicolumn{1}{c}{Sample} & \multicolumn{1}{c}{Country} & \multicolumn{1}{c}{Language}\\
\midrule
C2L1 & Rønneberg et al. (2022) & EyeWrite & Argumentative essays & 126 & 11.8 & 6th graders & Norway & Norwegian\\
CATO & Torrance et al. (2016) & EyeWrite & Expository texts & 52 & 16.9 & Secondary school students & Norway & Norwegian\\
GUNNEXP2 & Torrance and Ofstad (n.d) & EyeWrite & ????? & 45 & ???? & Undergraduate students & ???? Norway & ???? Norwegian\\
LIFT & Vandermeulen, Steendam, et al. (2020) & InputLog & Synthesis & 658 & 17 & Secondary school students & The Netherlands & Dutch\\
PLanTra & Rossetti and Van Waes (2022b) & InputLog & Text simplification & 47 & 23 & Master students & Belgium & English (L2)\\
SPL2 & Torrance et al. (n.d.) & CyWrite & Argumentative essays & 39 & 20.6 & Undergraduate students & USA & English (L1) / Spanish (L2)\\
\bottomrule
\end{tabular}

@@

\end{landscape}

\hypertarget{data-extraction}{%
\subsection{Data extraction}\label{data-extraction}}

From the keystroke data we extracted transition durations between adjacent keys at location that were found, by previous research, to be psycholinguistically meaningful (Chukharev-Hudilainen et al., 2019; De Smet et al., 2018; Torrance, Rønneberg, et al., 2016) and are detailed in Table \ref{tab:keyloc}. In particular we analysed the key-transitions that resulted in the insertion of a character that started a new sentence as before-sentence transition; transitions that started a new word other than those at the beginning of a sentence were treated as before-word transition; all transitions within a word but not the key-transition between the last letter of a word and the subsequent space or punctuation mark were treated as within-word transitions. At before-sentence locations, IKIs were timed to the shift keypress that resulted in the capitalization of the first key for most data sets (CATO, C2L1, SPL2, GUNNEXP2) but included the transition to the following sentence-initial letters in some data sets (PLanTra, LIFT); we return to this difference in the Results section. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions and were removed from the analysis. Also we removed transitions that were followed by an editing operation.

@@

\begin{tabular}{p{3cm}p{5cm}p{6cm}}
\toprule
Transition type & \multicolumn{1}{c}{Description} & \multicolumn{1}{c}{Example}\\
\midrule
Within word & Transitions between any letter & T$^{\wedge}$h$^{\wedge}$e c$^{\wedge}$a$^{\wedge}$t m$^{\wedge}$e$^{\wedge}$o$^{\wedge}$w$^{\wedge}$e$^{\wedge}$d. T$^{\wedge}$h$^{\wedge}$a$^{\wedge}$t[bsp][bsp]e$^{\wedge}$n i$^{\wedge}$t s$^{\wedge}$l$^{\wedge}$e$^{\wedge}$p$^{\wedge}$t.\\
Below word & Keypress after space followed by any letter & The $^{\wedge}$cat $^{\wedge}$meowed. That[bsp][bsp]en $^{\wedge}$it $^{\wedge}$slept.\\
Before sentence & Keypress following a space preceding any letter & The cat meowed. $^{\wedge}$That[bsp][bsp]en it slept.\\
\bottomrule
\addlinespace
\end{tabular}

@@

We removed participants that did not complete all conditions in studies with within-participant factors (reducing the number of participants to 343 for LIFT data, and 41 participants for PLanTra data). We removed participants that produced less than 10 sentences (LIFT: 109 participants; PLanTra: 3 participants; SPL2: 1 participant). We further removed keystroke intervals that were too short to represent a behavioural response (\(\le\) 50 msecs) or too long (\(\ge\) 30 secs) and possibly involving other task-unrelated activities; percentages can be found in Table \ref{tab:datareduction}. From the remaining data we randomly sampled a maximum of 100 observations per participant, condition, and transition location (when more than 100 were available). This was done to reduce the computation time of the Bayesian models require to complete sampling. For the LIFT data set we reduced the number of participants to 100 because the total sample was substantially larger than for the other datasets. Because the LIFT data set included the largest number of writing tasks, we sampled 50 observations per condition, location and participant which would otherwise exceed the computational resources available to us. The number of keystroke data used in the analysis can be found in Table \ref{tab:datareduction}.

@@

\caption{\label{tab:datareduction}Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data included by transition location (`--` if all data were used). Standard error in parentheses.}
